\chapter{Reproducing the Experiments}
\label{sec:reproducing_experiments}

The source and input models for evaluation of this project can be downloaded from \url{https://github.com/epsilonlabs/emf-cbp} and \url{https://drive.google.com/open?id=1W_DTOloflZt6aYvObDCTfcnSMqhiRJY1}.

All of the experiments of this research are developed and executed with the following specification Eclipe 4.6.3 (Neon.3) 32-bit and Java Development Kit 1.8. Some important libraries that are used in the source are BPMN2 Metamodel	1.3.0, Epsilon 1.4.0, EMF 2.12.0, UML2 Extender SDK	5.2.3, NeoEMF 1.0.2, 
EMF Store 1.9.0, EMF Compare 3.3.2, and MoDisco 1.0.1. These libraries can be installed as plugins on Eclipse. Other jar dependencies are Apache commons-io-2.6.jar for file operation utility and Java Specification Request jsr305-3.0.1.jar that is required by NeoEMF.

\begin{itemize}
  \item \textbf{Chapter 5}. The test code to run the evaluation of Chapter 5 is in file ECMFATest3.java. This file is located in package org.eclipse.epsilon.cbp.state2change.test under project
  org.eclipse.epsilon.cbp.state2change.
  \item \textbf{Chapter 6}. HybridXMITest.java, NeoEMFTest.java, HybridNeoEMFTest.java, and CBPTest.java are the test files for the evaluation of Chapter 6. They are located in package org.eclipse.epsilon.cbp.hybrid.test in project org.eclipse.epsilon. cbp.hybrid.test.
  \item \textbf{Chapter 7}. The test code to run the evaluation of Chapter 7 is in file CBPComparisonTest.java. This file is in package org.eclipse.epsilon.cbp.comparison. test under project
  org.eclipse.epsilon.cbp.comparison.
  \item \textbf{Chapter 8}. The test code to run the evaluation of Chapter 8 is in file CBPConflictTest.java Application.java under package org.eclipse.epsilon.cbp.conflict.test in project org.eclipse.epsilon.cbp.comparison and file
  Application.java under package org.eclipse.epsilon.cbp.comparison.emfstore.test in project org.eclipse. epsilon.cbp.comparison.emfstore. 
\end{itemize}